RationalGRL: A Framework for Argumentation and Goal Modeling

### 2018-05-07 discussion.tex
small change
rebuttal and small changes
updates

\paragraph{Specific Domains}
The argumentation schemes and critical questions presented in this paper are focused only on the core goal modeling process, that is, the discussion about the goals and functionality of an information system. In the RE process, there are many more domain specific discussions that also take place, such as discussions involving expert opinions \cite{murukannaiah2015}, discussions about security requirements \cite{haley2008security,yu2015automated,ionita2014argumentation}, architectural principles \cite{marosin-etal:caise2016}, legal compliance \cite{Ghanavati2013}, and so forth. From a knowledge engineering perspective, including the argument schemes and critical questions associated with these domains in the RationalGRL framework is a time-consuming task. However, from a formal perspective adding new schemes and questions is easy: as already discussed by Bex \emph{et al.}~\cite{bexEtal2003}, critical questions for argument schemes will always lead to either new information being introduced, new information replacing old information, or new information attacking old information. This corresponds to the \textsf{INTRO}, \textsf{REPLACE} and \textsf{DISABLE} operations in the RationalGRL framework, which have been formalized in Algorithms 3-9 in Section~\ref{sect:algorithms}, hence it would be suitable for other types of schemes and questions as well.

@@

\paragraph{Tool support}
The current tool is a prototype that implements the RationalGRL framework in a fairly simple way. That is, by offering the possibility to argue and ask critical questions about a goal model, and then export this goal model for further analysis in jUCMNav. Integration of RationalGRL elements into jUCMNav, or at least creating an import function to import jUCMNav goal models into the tool would allow us to tap into the large amount of research and development that is being carried out with jUCMNav. Furthermore, the tool could be expanded to include the new features described above, such as new argument schemes and reasoning with preferences. In addition to such extensions, it is interesting to think about new possibilities. One idea is to use the tool for collaborative on-line goal modeling, similar to Github: since the renaming (i.e. replacement) and deletion (i.e. disabling) of elements are all logged, it is easy for a stakeholder to continue working on a model that was made by another stakeholder. Furthermore, critical questions can be included for other stakeholders to answer at a later date. As a formal underpinning of this asynchronous communication between users, it would make sense to capture requirements engineering and software design processes as dialogs between parties~\cite{finkelstein1989multiparty,BlackEtal2013}, which are a natural fit with the question-answer format employed in the RationalGRL framework.

\paragraph{Empirical validation}
We have performed a case study and found more than 200 incidences of arguments and questions. However, what is still lacking is further empirical investigation of whether, and how, the tools and techniques provided by the RationalGRL framework really improve early phase requirements engineering. The complexity of the domain makes focused experiments difficult, but not impossible. For example, Murakannaiah \emph{et al.}~\cite{murukannaiah2015} provided a realistic but bounded problem for their subjects to solve, and similar experiments are imaginable to test RationalGRL: have two groups solve a goal modeling problem, one using only GRL and one using RationalGRL, and examine how they perform. Naturally, the type of problem will greatly influence the outcome. For a simple problem which two people have to solve in an hour, RationalGRL will most likely be of little benefit. However, a more complex problem on which multiple persons have to work asynchronously might benefit from the extra tools offered by RationalGRL. Furthermore, the outcome also depends on which part of the set of tools and techniques offered by RationalGRL is tested: the development process, the tool, or a list of critical questions to serve as reminders during the goal modeling process (cf. \cite{SchriekEtal2016}).

@@

One of our main goals is to provide means for RE practitioners to capture the underlying arguments of goal models by using the RationalGRL framework (\textbf{requirement 4}). To this end, we propose a development process, which consists of developing goal models and posing arguments based on schemes in an integrated way. Finally, we have implemented the RationalGRL tool, a web-based prototype\footnote{\url{http://www.rationalgrl.com}}, for modeling argument schemes and critical questions and for reasoning about goal models with respect to them (\textbf{requirement 5}).

>>>>

\paragraph{Specific domains}
The argumentation schemes and critical questions presented in this paper are focused only on the core goal modeling process, that is, the discussion about the goals and functionality of an information system. In the RE process, there are many more domain specific discussions that also take place, such as discussions involving expert opinions \cite{murukannaiah2015}, discussions about security requirements \cite{haley2008security,yu2015automated,ionita2014argumentation}, architectural principles \cite{marosin-etal:caise2016}, legal compliance \cite{Ghanavati2013}, and so forth. From a knowledge engineering perspective, including the argument schemes and critical questions associated with these domains in the RationalGRL framework is a time-consuming task. However, from a formal perspective adding new schemes and questions is easy: as already discussed by Bex \emph{et al.}~\cite{bexEtal2003}, critical questions for argument schemes will always lead to either new information being introduced, new information replacing old information, or new information attacking old information. This corresponds to the \textsf{INTRO}, \textsf{REPLACE} and \textsf{DISABLE} operations in the RationalGRL framework, which have been formalized in Algorithms 3-9 in Section~\ref{sect:algorithms}, hence it would be suitable for other types of schemes and questions as well.

@@

\paragraph{Process and tool support}
The current tool is a prototype that implements the RationalGRL framework in a fairly simple way. That is, by offering the possibility to argue and ask critical questions about a goal model, and then export this goal model for further analysis in jUCMNav. Integration of RationalGRL elements into jUCMNav, or at least creating an import function to import jUCMNav goal models into the tool would allow us to tap into the large amount of research and development that is being carried out with jUCMNav. Furthermore, the tool could be expanded to include the new features described above, such as new argument schemes and reasoning with preferences.

Additionally, from the user evaluation of the tool it followed that in order to implement the tool in realistic settings, other ways may have to be found to incorporate arguments in the goal modeling process. One idea is to use the tool for collaborative on-line goal modeling, similar to Github: since the renaming (i.e. replacement) and deletion (i.e. disabling) of elements are all logged, it is easy for a stakeholder to continue working on a model that was made by another stakeholder. Furthermore, critical questions can be included for other stakeholders to answer at a later date. As a formal underpinning of this asynchronous communication between users, it would make sense to capture requirements engineering and software design processes as dialogs between parties~\cite{finkelstein1989multiparty,BlackEtal2013}, which are a natural fit with the question-answer format employed in the RationalGRL framework. Finally, it would make sense to further develop the RationalGRL Development Process. One idea here is to have stakeholders play a card game (similar to the well-known planning poker) with cards that contain critical, reflective questions (cf. \cite{SchriekEtal2016})

\paragraph{Empirical validation}
We have performed a case study and found more than 200 incidences of arguments and questions. \rood{Zeg ook dat we user study gedaan hebben, en noem het idee om in de toekomst een "industrial-strength case-study" te doen}. However, what is still lacking is further empirical investigation of whether, and how, the tools and techniques provided by the RationalGRL framework really improve early phase requirements engineering. The complexity of the domain makes focused experiments difficult, but not impossible. For example, Murakannaiah \emph{et al.}~\cite{murukannaiah2015} provided a realistic but bounded problem for their subjects to solve, and similar experiments are imaginable to test RationalGRL: have two groups solve a goal modeling problem, one using only GRL and one using RationalGRL, and examine how they perform. Naturally, the type of problem will greatly influence the outcome. For a simple problem which two people have to solve in an hour, RationalGRL will most likely be of little benefit. However, a more complex problem on which multiple persons have to work asynchronously might benefit from the extra tools offered by RationalGRL. Furthermore, the outcome also depends on which part of the set of tools and techniques offered by RationalGRL is tested: the development process, the tool, or a list of critical questions to serve as reminders during the goal modeling process (cf. \cite{SchriekEtal2016}).

@@

One of our main goals is to provide means for RE practitioners to capture the underlying arguments of goal models by using the RationalGRL framework (\textbf{requirement 4}). To this end, we propose a development process, which consists of developing goal models and posing arguments based on schemes in an integrated way.

Finally, we have implemented the RationalGRL tool, a web-based prototype\footnote{\url{http://www.rationalgrl.com}}, for modeling argument schemes and critical questions and for reasoning about goal models with respect to them (\textbf{requirement 5}). We performed a small empirical evaluation with 16 users. The results indicate that while the cognitive overhead of RationalGRL (or goal modeling in general) is high, combining goal modeling with argumentation is perceived as useful. Furthermore, using a formal semantics to compute which arguments and ultimately goal models are acceptable given the stakeholders' opinions valid argument was perceived as very useful.

### 2018-05-07 introduction.tex
rebuttal and small changes
updates

Inspired by the work on practical reasoning from Artificial Intelligence, most notably Atkinson and Bench-Capon~\cite{atkinson2007}, we have developed a list of argument schemes that can be used to analyze and guide stakeholders' discussions about goal models. This list of argument schemes is based on an extensive case study in which we analyzed a set of transcripts containing more than 4 hours of discussions among designers of a traffic simulator information system.

In order to specify clearly in what way RationalGRL extends GRL, we develop a metamodel, which specifies the traceability links between the arguments based on the schemes and the GRL models. In addition to this meta-model, we provide formal semantics for RationalGRL by formalizing the GRL language in propositional logic and rendering arguments about a GRL model as a formal argumentation framework to which we can apply argumentation semantics~\cite{Dung1995}. We, then, formally capture the links between argumentation and goal modeling as a set of algorithms for applying argument schemes and critical questions about goal models.

In order to support practitioners in using the RationalGRL framework, we propose the RationalGRL development process, which explains steps for  developing goal models and posing arguments based on schemes in an integrated way. We implement a web-based tool for the RationalGRL framework as well, which works on any modern browser and is developed in Javascript.\footnote{The tool and source code can be found at \rationalgrlurl.} The tool is open-source, and designed such that new argument schemes and critical questions can be added easily.

The rest of this article is organized as follows. Section~\ref{sect:background} introduces our running example and briefly discusses the basics of GRL and argumentation about goals using argument schemes. Section~\ref{sect:gmas} contains the case study and an explanation of how we obtained an initial set of argument schemes and critical questions by coding transcripts from discussions about an information system. Section~\ref{sect:overview} provides an overview of the RationalGRL framework and the RationalGRL metamodel, and examples from the case study that illustrate the framework and its language. In Section~\ref{sect:formalframework}, we provide formal semantics for GRL and RationalGRL, show how RationalGRL models can be translated to GRL models and vice versa, and we develop various algorithms that change a RationalGRL model according to an argument scheme or a critical question. Section~\ref{sect:methodology+tool} discusses the RationalGRL development process and explains various features of the tool we developed. Finally, Section~\ref{sect:discussion} covers the related work, the future work, and the conclusion.

>>>>

Inspired by the work on practical reasoning from Artificial Intelligence, most notably Atkinson and Bench-Capon~\cite{atkinson2007}, we have developed a list of argument schemes that can be used to analyze and guide stakeholders' discussions about goal models. This list of argument schemes is based on an extensive case study in which we analyzed a set of transcripts containing more than 4 hours of discussions among designers of a traffic simulator information system (requirement 1).

In order to specify clearly in what way RationalGRL extends GRL, we develop a metamodel, which specifies the traceability links between the arguments based on the schemes and the GRL models (requirement 2). In addition to this meta-model, we provide formal semantics for RationalGRL by formalizing the GRL language in propositional logic and rendering arguments about a GRL model as a formal argumentation framework to which we can apply argumentation semantics~\cite{Dung1995}. We, then, formally capture the links between argumentation and goal modeling as a set of algorithms for applying argument schemes and critical questions about goal models (requirement 3).

In order to support practitioners in using the RationalGRL framework, we propose the RationalGRL development process, which explains steps for developing goal models and posing arguments based on schemes in an integrated way (requirement 4). We implement and briefly evaluate a web-based tool for the RationalGRL framework, which works on any modern browser and is developed in Javascript (requirement 5).\footnote{The tool and source code can be found at \rationalgrlurl.} The tool is open-source, and designed such that new argument schemes and critical questions can be added easily.

The rest of this article is organized as follows. Section~\ref{sect:background} introduces our running example and briefly discusses the basics of GRL and argumentation about goals using argument schemes. Section~\ref{sect:gmas} contains the case study and an explanation of how we obtained an initial set of argument schemes and critical questions by coding transcripts from discussions about an information system. Section~\ref{sect:overview} provides an overview of the RationalGRL framework and the RationalGRL metamodel, and examples from the case study that illustrate the framework and its language. In Section~\ref{sect:formalframework}, we provide formal semantics for GRL and RationalGRL, show how RationalGRL models can be translated to GRL models and vice versa, and we develop various algorithms that change a RationalGRL model according to an argument scheme or a critical question. Section~\ref{sect:methodology+tool} discusses the RationalGRL development process and explains various features of the tool we developed. Section~\ref{sect:validation} contains a brief validation of RationalGRL. We conducted a small user study (16 participants), where we asked the participants to model a small example in the RationalGRL, and to answer some questions on the usability and usefulness RationalGRL. Finally, Section~\ref{sect:discussion} covers related work, future work, and the conclusion.

### 2018-05-07 validation.tex
updates

Since RationalGRL is targeted at practitioners and not just for academic end users, we performed a brief user evaluation with 16 users. In this evaluation, we asked participants to rate and comment the usability of the RationalGRL tool. In this section we first present the results of the evaluation in three subsections, and then provide a more high-level analysis in the last subsection.

@@

The instructions for these three parts are contained in a single document that we sent out to the participants.\footnote{See \url{www.rationalgrl.com}, page ``Empirical Study''} At the end of the document we provide a link to the survey.\footnote{https://goo.gl/forms/fDSMUnAV20wy7kbY2} The instructions we provide to the users for task 2 can be found in Appendix~\ref{sect:survey}, and some example RationalGRL models the users created can be found in Appendix~\ref{sect:survey-screenshots}.

Two thirds of the participants had either a PhD or Master degree, 20\% a Bachelor degree, and one respondent responded with ``Other''. The participants turned out to have a relatively high level of experience with IT, with an average rating of 6 years, ranging from 0 to 25 years. They also judged themselves to be quite experienced with early-phase requirements engineering (average of rating 3.2 out of 5), but experience with goal modeling was less (rating 1.9 out of 5). Goal modeling languages the participants had experience with were GRL (2), i* (5), Tropos (1), and KAOS (1).

@@

We also asked the users three open questions related to usability. When asked about the strengths and weaknesses of RationalGRL, users where generally positive about RationalGRL's clear UI and the fact that arguments can be combined with goal models. For instance, one user stated that RationalGRL successfully ``tries to capture the rational behind the modeling process' and ``has a simple) way to compute the status of the arguments''. Users did found it challenging to decide when to use arguments, and to determine when an arguments attacks another argument. For instance, user mentioned ``it wasn't very easy because i didn't always know how to attack the elements'', and ``i dont think its necessary to include all the arguments because some of them are a bit trivial''. Other users were worried about the cognitive overload of the tool. One user mentioned ``I still believe in the value of arguments, but there should be less confusing ways to capture them''.

@@

When asked whether the users believe a more mature version of RationalGRL can be used to capture early-phase requirements in an actual software development project, most users responded positively. This was somewhat surprising, since our tool is still an early prototype. Users' responses include ``Yes, having the possibility to add arguments seems quite useful'', ``I think it can, but maybe try to find a way to combine it with more regular/mainstream requirements gathering such as user stories and customer journeys'', ``yes, because it creates a clear scope for the project, what the goals are'', and ``Generally I think its useful to explicitly document arguments of a discussion''. The worry that there may be a cognitive overhead did come back here as well: ``I think the overhead of inputting a (detailed) discussion in a structured manner into any system makes adoption difficult'', and ``the manual input is too complex and takes too much time. An automated process of parsing the conversation log would be much more helpful''.

@@

\subsection{Analysis}
Although our empirical evaluation is relatively small-scale, and we cannot draw hard conclusions from it, we do believe that we can extract some interesting observations. We list the three most important ones:
\begin{enumerate}
\item \emph{High cognitive overload.} A worry that came back often in our empirical evaluation is that, in its current form, RationalGRL has a relatively high cognitive overload. Goal modeling is by itself already a relatively high-effort activity, and the fact that we add more elements to the language does not improve this. This may mean that in future work we should focus more on the RationalGRL development process. It seems users do like to be guided during modeling, and making arguments explicit in this process seems useful, but it could be integrated into the process of goal modeling in a more natural way. Critical questions seem to be a natural fit for this. They play a key role in the RationalGRL development process, but in its current form they are relatively difficult to access (in the details pane).
\item \emph{Useful argumentation semantics.} Participants were enthusiastic about adding arguments to a goal modeling language with a clear formal underpinning, and they believed the argumentation semantics we use in the tool is very intuitive. This is a positive signal for our formal approach.
\item \emph{Lightweight tool.} Respondents were positive about the fact that it was very easy to get started with the tool. Thus it seems that having a simple web-interface for goal modeling languages is something that could be done more in the future. Most comments on the tool were the type of comments that are expected from a prototype implementation.
\end{enumerate}

>>>>

In addition to the empirical case study (Section~\ref{sect:gmas}), which forms the basis of the RationalGRL framework, we also performed a small user evaluation with 16 users. This evaluation concerned  mainly the use of the RationalGRL tool, but more general questions regarding the usefulness of arguments in goal modelling were also asked. We first present the details and results of the evaluation, and then provide a more high-level analysis in Section~\ref{sec:eval-an}.

@@

The instructions for these three parts are contained in a single document that we sent out to the participants.\footnote{See \url{www.rationalgrl.com}, page ``Empirical Study''} At the end of the document we provide a link to the survey.\footnote{https://goo.gl/forms/fDSMUnAV20wy7kbY2} The instructions we provide to the users for task 2 can be found in Appendix~\ref{sect:survey}.

Two thirds of the participants had either a PhD or Master degree, 20\% a Bachelor degree, and one respondent responded with ``Other''. The participants turned out to have a relatively high level of experience with IT, with an average rating of 6 years, ranging from 0 to 25 years. They also judged themselves to be quite experienced with early-phase requirements engineering (average of rating 3.2 out of 5), but experience with goal modeling was less (rating 1.9 out of 5, although two respondents use goal modeling languages almost weekly). Goal modeling languages the participants had experience with were GRL (2), i* (5), Tropos (1), and KAOS (1).

In task 2, users clearly made use of the extra features offered by the RationalGRL tool: they used the details pane more often than not (rating 3.2 out of 5) and on average produced 3 arguments from the short transcript. Some example RationalGRL models the users created can be found in Appendix~\ref{sect:survey-screenshots}.

@@

We also asked the users three open questions related to usability. When asked about the strengths and weaknesses of RationalGRL, users where generally positive about RationalGRL's clear UI and the fact that arguments can be combined with goal models. For instance, one user stated that RationalGRL successfully ``tries to capture the rationale behind the modeling process' and ``has a simple way to compute the status of the arguments''. Users did found it challenging to decide when to use arguments, and to determine when an arguments attacks another argument. For instance, user mentioned ``it wasn't very easy because I didn't always know how to attack the elements'', and ``I don't think its necessary to include all the arguments because some of them are a bit trivial''. Other users were worried about the cognitive overload of the tool. One user mentioned ``I still believe in the value of arguments, but there should be less confusing ways to capture them''.

@@

When asked whether the users believe a more mature version of RationalGRL can be used to capture early-phase requirements in an actual software development project, most users responded positively. This was somewhat surprising, since our tool is still an early prototype. Users' responses include ``Yes, having the possibility to add arguments seems quite useful'', ``I think it can, but maybe try to find a way to combine it with more regular/mainstream requirements gathering such as user stories and customer journeys'', ``yes, because it creates a clear scope for the project, what the goals are'', and ``Generally I think its useful to explicitly document arguments of a discussion''. The worry that the modeling process may involve too much cognitive overhead was mentioned again though: ``I think the overhead of inputting a (detailed) discussion in a structured manner into any system makes adoption difficult'', and ``the manual input is too complex and takes too much time. An automated process of parsing the conversation log would be much more helpful''.

@@

\subsection{Analysis}\label{sec:eval-an}
Although our empirical evaluation is relatively small-scale, and we cannot draw hard conclusions from it, we do believe that we can extract some interesting observations. We list the three most important ones:
\begin{enumerate}
\item \emph{High cognitive overhead.} A worry that came back often in our empirical evaluation is that, in its current form, RationalGRL has a relatively high cognitive overhead. Goal modeling is by itself already a cognitively high-effort activity, and the fact that we add more elements to the language does not improve this. This may mean that in future work we should focus more on the RationalGRL development process\footnote{Note that in order to keep the study simple for the users, we did not explicitly ask the respondents to follow the development process from Section~\ref{sect:methodology}}. Users like to be guided during modeling, and making arguments explicit in the modelling process was indicated as useful, but argumentation should be integrated into the \emph{process} of goal modeling more than in the goal models themselves. Critical questions seem to be a natural fit for this, as they play a key role in the RationalGRL development process, but in its current form they are only mentioned in the details pane of the tool.
\item \emph{Useful argumentation semantics.} Participants were enthusiastic about adding arguments to a goal modeling language with a clear formal underpinning, and they believed the argumentation semantics we use in the tool is very intuitive. This is a positive signal for our formal approach.
\item \emph{Lightweight tool.} Respondents were positive about the fact that it was very easy to get started with the tool. Thus it seems that having a simple web-interface for goal modeling languages is something that could be done more in the future. Most comments on the tool were the type of comments that are expected from a prototype implementation.
\end{enumerate}
